---
title: "Rerun-outlier-removal"
author: "Allison Davis Connelly & Sophia McKelvey"
date: "2026-04-23"
output: 
  html_document:
    toc: TRUE
    toc_depth: 5
    toc_float:
      collapsed: FALSE
      
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# **PURPOSE:**

When exploring our data, we noticed a few outliers and wanted to investigate the how much of an effect they had on our results. With a small sample size, as we have in this experiment, it is difficult to tell if they are true outliers or within a range of natural variation. Additionally, since all identified outliers are females, it is possible that reproductive status (a factor we did not control for, although all females tested were not visibly pregnant) is altering the thermal tolerance. Therefore, these analyses are for exploratory purposes only, and we will not be removing these outliers from our final reported data set. 

The analysis below is a repeat of the analyses found in *Final_temperature-analysis.Rmd* within the **Thermal Tolerance Analysis** folder. To run this current file, the *Temp ID - Completed Data.csv* from the **Thermal Tolerance Analysis** folder is required. 


***
***
***


# **Data**

All analyses were conducted using R version 4.4.2 

## Load required libraries

Remove the '#' prior to `install.packages` if it is the first time installing library. 

```{r}
#install.packages("curl")             #v5.2.1
library(curl)

#install.packages("dplyr")            #v1.1.4
library(dplyr)

#install.packages("summarytools")     #v1.1.4
library(summarytools)

#install.packages("gt")               #v0.11.1
library(gt)

#install.packages("tidyr")            #v1.3.1
library(tidyr)
  
#install.packages("MuMIn)              #v1.48.4
library(MuMIn)

#install.packages("lmerTest")         #v3.1.3
library(lmtest)

#install.packages("sandwich")         #v3.1-1
library(sandwich)

#install.packages("ggplot2")          #v3.5.1
library(ggplot2)

#install.packages("ggpubr")          #v0.6.2
library(ggpubr)

#install.packages("DHARMa")           #v0.4.7
library(DHARMa)

#install.packages("emmeans")          #v1.10.5
library(emmeans)

#install.packages("car")              #v3.1-2
library(car)

#install.packages("patchwork")        #v1.3.2
library(patchwork)


```

## Read in data

```{r}

# OPTION 1: download 'Temp ID-Completed Data.csv' into working directory 
rawTempData <- read.csv("Temp ID - Completed Data.csv", header = TRUE, sep = ",", stringsAsFactors = TRUE)
head(rawTempData)


# OPTION 2: load from Github repository
rawTempData <- curl("https://raw.githubusercontent.com/sophia-m0823/TestRStudio/main/Temp%20ID%20-%20Completed%20Data.csv")
rawTempData <- read.csv(rawTempData, header = TRUE, sep = ",", stringsAsFactors = TRUE)
head(rawTempData)

#add common names 
rawTempData$Species <- factor(rawTempData$Species,
                        levels = c("X. nigrensis", "P. latipinna", "G. affinis"),
                        labels = c("Swordtail", "Sailfin", "Mosquitofish"))


```

## Identifying outliers

```{r}

#calculate outliers for both CTmax and CTmin by species
df_flagged <- rawTempData %>%
  group_by(Species) %>%
  mutate(
    q1_max = quantile(CTmax.C, 0.25, na.rm = TRUE),
    q3_max = quantile(CTmax.C, 0.75, na.rm = TRUE),
    iqr_max = q3_max - q1_max,
    q1_min = quantile(CTmin.C, 0.25, na.rm = TRUE),
    q3_min = quantile(CTmin.C, 0.75, na.rm = TRUE),
    iqr_min = q3_min - q1_min,
    is_outlier_max = CTmax.C < (q1_max - 1.5 * iqr_max) | CTmax.C > (q3_max + 1.5 * iqr_max),
    is_outlier_min = CTmin.C < (q1_min - 1.5 * iqr_min) | CTmin.C > (q3_min + 1.5 * iqr_min)) %>%
  ungroup()

#visualize CTmin outliers
out.min <- ggplot(df_flagged, aes(x = CTmin.C, fill = is_outlier_min)) +
  geom_histogram(binwidth = 0.5, color = "white") +
  facet_wrap(~Species, scales = "free_y") +
  scale_fill_manual(values = c("FALSE" = "grey70", "TRUE" = "#d95f02"),
                    name = "Data Status",
                    labels = c("Normal", "Outlier")) +
  labs(title = "CTmin",
       x = "Critical Thermal Minimum (°C)",
       y = "Frequency") +
  theme_minimal()
out.min #outliers for sailfin

#visualize CTmax outliers
out.max <- ggplot(df_flagged, aes(x = CTmax.C, fill = is_outlier_max)) +
  geom_histogram(binwidth = 0.5, color = "white") +
  facet_wrap(~Species, scales = "free_y") +
  scale_fill_manual(values = c("FALSE" = "grey70", "TRUE" = "#d95f02"),
                    name = "Data Status",
                    labels = c("Normal", "Outlier")) +
  labs(title = "CTmax",
       x = "Critical Thermal Minimum (°C)",
       y = "Frequency") +
  theme_minimal()
out.max #outliers for swordtail and mosquitofish

((out.min/out.max) + plot_annotation(tag_levels = "a") + plot_layout(guides = "collect"))

#filter for rows where either CTmax OR CTmin is an outlier
outlier_individuals <- df_flagged %>%
  filter(is_outlier_max | is_outlier_min) %>%
  select(Species, Sex, Name, CTmax.C, CTmin.C, is_outlier_max, is_outlier_min)
outlier_individuals #all females

#filter out outliers from df
O.rawTempData <- rawTempData %>%
  filter(!Name %in% c("Toast", "Cherrios", "Sesame", "Ellie", "Rory", "Loreli", "Lane"))



```


## Summery stats 

Trial stat changes:

  - Species: CTmax OLD TEMP -> NEW TEMP; CTmin OLD TEMP -> NEW TEMP
  - Swordtail: CTmax 35.27 -> 35.53; CTmin 9.86 -> 9.88
  - Sailfin: CTmax 38.44 -> 38.38; CTmin 6.88 -> 6.98
  - Mosquitofish: CTmax 37.58 -> 38.18; CTmin 4.95 -> 4.56

```{r}
#remove unnecessary columns
O.datTot <- O.rawTempData[,-c(5:7, 10)]

#create long format df to easily summarize later
O.dat.long <- O.datTot %>%
  pivot_longer(
    cols = c(CTmax.C, CTmin.C),
    names_to = "Trial",
    values_to = "Value") %>%
  mutate(
    Trial = dplyr::recode(
      Trial,
      "CTmax.C" = "CTmax",
      "CTmin.C" = "CTmin"))

#gather summary stats
O.summ.stats <- O.dat.long %>%
  group_by(Species, Trial) %>%
  summarize(
    mean = mean(Value, na.rm = TRUE),
    sd   = sd(Value, na.rm = TRUE),
    min  = min(Value, na.rm = TRUE),
    med  = median(Value, na.rm = TRUE),
    max  = max(Value, na.rm = TRUE),
    .groups = "drop")

#get n for each species and sex
O.sex.df <- O.datTot %>%
  count(Species, Sex) %>%
  tidyr::pivot_wider(
    names_from  = Sex,
    values_from = n,
    values_fill = 0)

#final table
O.summ.stats.all <- O.summ.stats %>%
  left_join(O.sex.df, by = "Species") %>%
  mutate(
    mean_sd = sprintf("%.2f (%.2f)", mean, sd))

O.summ.stats.all


```


Size stat changes: since only females were removed, we will only report female changes

  - Species: OLD SIZE -> NEW SIZE
  - Swordtail: 23.1 -> 23.0
  - Sailfin: 32.9 -> 32.8
  - Mosquitofish: 30.2 -> 30.0
  - Overall female size change: 28.6 -> 28.5


```{r}
# Average lengths by species & sex

#gather summary stats
O.sizes <- O.datTot %>%
  group_by(Species, Sex) %>%
  summarise(
    avg.len = mean(Size.mm),
    st.dev = sd(Size.mm),
    .groups = "drop")

#final table
 O.sizes.fmt <- O.sizes %>%
   mutate(
    size.display = sprintf("%.1f (%.2f)", avg.len, st.dev)) %>%
  select(Species, Sex, size.display) %>%
  pivot_wider(
    names_from = Sex,
    values_from = size.display)
O.sizes.fmt

### boxplot

#create more intuitive sex categories for swordtails
O.dat.plot <- O.datTot %>%
  mutate(
    O.Sex.plot = case_when(
      Species == "Swordtail" & Sex == "male" & Size.cat == "small" ~ "Male (small)",
      Species == "Swordtail" & Sex == "male" & Size.cat == "large" ~ "Male (large)",
      Sex == "male"                                             ~ "Male",
      Sex == "female"                                           ~ "Female"))


#final plot
ggplot(O.dat.plot, aes(x = O.Sex.plot, y = Size.mm)) +
  geom_rect(
    data = subset(O.datTot, Species == "Sailfin"),
    inherit.aes = FALSE,
    xmin = -Inf, xmax = Inf,
    ymin = -Inf, ymax = Inf,
    fill = "grey95") +
  geom_boxplot() +
  geom_point(color= "#8D8680", size=2, alpha=0.5) +
  facet_wrap(~ Species, scales = "free_x") +
  scale_x_discrete(
  labels = function(x) {
    ifelse(x %in% c("Male (small)", "Male (large)"),
           gsub(" ", "\n", x),
           x)}) +
  theme_classic() + 
  xlab("") +
  ylab("Length (mm)")
  



# Average lengths by sex

#gather summary stats
O.sizes2 <- O.datTot %>%
  group_by(Sex) %>%
  summarise(
    avg.len = mean(Size.mm),
    st.dev = sd(Size.mm),
    .groups = "drop")

#final table
 O.sizes.fmt2 <- O.sizes2 %>%
   mutate(
    size.display = sprintf("%.1f (%.2f)", avg.len, st.dev)) %>%
  select(Sex, size.display) %>%
  pivot_wider(
    names_from = Sex,
    values_from = size.display)
 O.sizes.fmt2
 

### boxplot

#final plot
ggplot(O.datTot, aes(x = Sex, y = Size.mm)) +
  geom_boxplot() +
  geom_point(color= "#8D8680", size=2, alpha=0.5) +
  theme_classic() +
  xlab("") +
  ylab("Length (mm)") +
  scale_x_discrete(labels=c('Female', 'Male'))
 
 
```


# **Analysis**

## **CTmax**

### Model Selection


**Results:** The model containing just species as a dependent variable (O.mod1f) is still the most supported in our model selection (ΔAICc=0.0; weight=0.539). The model with both species and sex still has a delta value less than 2 (ΔAICc=1.25; weight=0.288), so we will still keep this model for the purpose of illustrating the limited effect of sex. 

```{r}
# All models
O.mod1 <- lm(CTmax.C ~ Species*Sex + Size.mm, data = O.datTot) #full model with interaction effect
O.mod1b <- lm(CTmax.C ~ Species + Sex + Size.mm, data = O.datTot) #full model, no interaction
O.mod1c <- lm(CTmax.C ~ Species + Sex, data = O.datTot) #no size effect
O.mod1d <- lm(CTmax.C ~ Sex + Size.mm, data = O.datTot) #no species effect
O.mod1e <- lm(CTmax.C ~ Sex, data = O.datTot) #sex only effect
O.mod1f <- lm(CTmax.C ~ Species, data = O.datTot) #species only effect
O.mod1g <- lm(CTmax.C ~ Size.mm, data = O.datTot) #size only effect

# Model selection
model.sel(O.mod1, O.mod1b, O.mod1c, O.mod1d, O.mod1e, O.mod1f, O.mod1g)

  ### O.mod1f (species only effect) is the most supported model (ΔAICc=0; weight=0.539)
  ### O.mod1c (species and sex, no size) also has a ΔAICc < 2 (ΔAICc=1.3; weight=0.288), so will keep to illustrate the limited effects of sex, but O.mod1f is best.

```

### Residual checks


**Results:** Both the species only model (O.mod1f, most supported) and the sex and species model (O.mod1c)  NO LONGER show a significant deviations from normality. Robust standard errors are no longer needed.

```{r}

# Species only model
simulationOutput <- simulateResiduals(O.mod1f)

plot(simulationOutput)

plotResiduals(simulationOutput)
plotResiduals(simulationOutput, form = O.rawTempData$Species)
   ### All good!

# Species and sex model
simulationOutput2 <- simulateResiduals(O.mod1c)

plot(simulationOutput2)

plotResiduals(simulationOutput2)
plotResiduals(simulationOutput2, form = O.rawTempData$Species)
plotResiduals(simulationOutput2, form = O.rawTempData$Sex)

  ### All good!


```


### Model Summary

**Results:** There is a significant effect of species for CTmax (P<0.001). Pairwise comparisons show a significant difference between swordtails and sailfin (p<0.001) and mosquitofish (p<0.001), but no difference between sailfin & mosquitofish (p=0.620). Sex was again confirmed as a non-significant predictor in the model (estimate= -0.159, SE=0.158, p=0.318). *These results mirror full-data results.*

```{r}

# SPECIES ONLY MODEL

  ## Regular check (not robust)
summary(O.mod1f)

  ## Overall species effect
Anova(O.mod1f, type="II")

 ## EMM & Pairwise comparisons

PW.sp.max <- emmeans(O.mod1f, "Species")
summary(PW.sp.max) #EMM
pairs(PW.sp.max) #Pairwise results
      
  ## Plotting EMM

PW.df <- as.data.frame(PW.sp.max)

ggplot(PW.df, aes(x=Species, y=emmean)) +
  geom_point(size=2) +
  geom_errorbar(aes(ymin=lower.CL, ymax=upper.CL), width=0.1) +
  scale_y_continuous(breaks = scales::pretty_breaks(n = 10)) +
  ylab("Predicted mean temperature (°C) \n") +
  xlab("\n Species") +
  theme_classic()

# SPECIES & SEX MODEL

summary(O.mod1c)


```

## **CTmin**

### Model Selection


**Results:** The species only model (O.mod2f) was again the most supported in our model selection (ΔAICc=0.0; weight=0.416). However, the full model with no interactions (O.mod2b) and the sex and species model (O.mod2c) also have a ΔAICc <2 (full model: ΔAICc= 0.8, weight= 0.279; spp+sex: ΔAICc = 1.07, weight=0.243). We will keep these two models to illustrate the effect of size and sex on our data, but maintain the species only model (O.mod2f) as our most supported model.

```{r}
# All models
O.mod2 <- lm(CTmin.C ~ Species*Sex + Size.mm, data = O.datTot) #full model with interaction effect
O.mod2b <- lm(CTmin.C ~ Species + Sex + Size.mm, data = O.datTot) #full model, no interaction
O.mod2c <- lm(CTmin.C ~ Species + Sex, data = O.datTot) #no size effect
O.mod2d <- lm(CTmin.C ~ Sex + Size.mm, data = O.datTot) #no species effect
O.mod2e <- lm(CTmin.C ~ Sex, data = O.datTot) #sex only effect
O.mod2f <- lm(CTmin.C ~ Species, data = O.datTot) #species only effect
O.mod2g <- lm(CTmin.C ~ Size.mm, data = O.datTot) #size only effect

# Model selection
model.sel(O.mod2, O.mod2b, O.mod2c, O.mod2d, O.mod2e, O.mod2f, O.mod2g)

  ### O.mod2f (species only) is the most supported model (ΔAICc=0; weight=0.416)
  ### O.mod2b (full model, no interactions) also has a ΔAICc < 2 (ΔAICc=0.80; weight=0.279)
  ### O.mod2c (species + sex) also has a ΔAICc < 2 (ΔAICc=1.07; weight=0.243)
  ### Will check out the summaries below to see the degree sex and size impact the model; maintain O.mod2f as most supported

```

### Residual checks

**Results:** Both the species only model (O.mod2f) and the sex and species model (O.mod2c) are heteroscedastic. This is driven by species-level differences in variation. We will use robust standard errors to assess the significance of the effects despite this heteroscedasticity. The full model with no interaction (O.mod2b) passes residual checks, so no robust standard errors are needed.

```{r}

# Species only model
simulationOutput <- simulateResiduals(O.mod2f)

plot(simulationOutput)

plotResiduals(simulationOutput)
plotResiduals(simulationOutput, form = O.rawTempData$Species)

  ### Failed Levene's test driven by species-level variation differences



# full model, no interactions
simulationOutput <- simulateResiduals(O.mod2b)

plot(simulationOutput)

plotResiduals(simulationOutput)
plotResiduals(simulationOutput, form = O.rawTempData$Species)

  ### All good!



# Species and sex model
simulationOutput <- simulateResiduals(O.mod2c)

plot(simulationOutput)

plotResiduals(simulationOutput)
plotResiduals(simulationOutput, form = O.rawTempData$Species)
plotResiduals(simulationOutput, form = O.rawTempData$Sex)

  ### Fine in overall plots, but by species shows a failure in Levene's test


```


### Model Summary

**Results:** There is still a significant effect of species for CTmin (P<0.001). Pairwise comparisons show all species as significantly different: swordtail-sailfin= p<0.001; swordtail-mosquitofish=p<0.001; sailfin-mosquitofish= p=0.001). Sex remained a non-significant predictor in both the full model with no interactions (p=0.28) and the sex + species model (p=0.28). Additionally, size is not a significant predictor in the full model with no interactions (p=0.07). *These results mirror full-data results.*

```{r}
# SPECIES ONLY MODEL

## Robust variance-covariance matrix
robVCV.sp <- vcovHC(O.mod2f, type= "HC3")

 ## Overall species effect
Anova(O.mod2f, vcov = robVCV.sp, type="II")

 ## EMM & Pairwise comparisons

PW.sp.min <- emmeans(O.mod2f, "Species", vcov. = robVCV.sp)
summary(PW.sp.min) #EMM
pairs(PW.sp.min) #Pairwise results
      
 ## Plotting EMM

PW.df <- as.data.frame(PW.sp.max)

ggplot(PW.df, aes(x=Species, y=emmean)) +
  geom_point(size=2) +
  geom_errorbar(aes(ymin=lower.CL, ymax=upper.CL), width=0.1) +
  scale_y_continuous(breaks = scales::pretty_breaks(n = 10)) +
  ylab("Predicted mean temperature (°C) \n") +
  xlab("\n Species") +
  theme_classic()



# FULL MODEL NO INTERACTIONS
coeftest(O.mod2b, vcov. = vcovHC(O.mod2b, type= "HC3")) #sex not sig (p=0.277), size not sig (p=0.070)



# SPECIES & SEX MODEL
coeftest(O.mod2c, vcov. = vcovHC(O.mod2c, type= "HC3")) #sex not sig (p=0.283)




```

## **Across mating strategies**

The second aim in our analysis is to compare thermal tolerance across male mating strategies within one species (Panuco swordtail). We used swordtail males of two categories: large, courting males (standard length >32mm) or small, coercive males (standard length <25mm with clearly defined gonopodium). If preexisting sex-based differences (e.g. bright coloration, elongated fins, courtship) affect thermal tolerance, we predict large males to have a lower thermal tolerance than small males, and small males to mirror the thermal tolerance of females.  

No males were removed in our outlier removal, but three female swordtails were removed. 

### Models

Since male mating strategy is associated with size, adding size to our models would be redundant. Just to confirm, we will run a correlation test (kruskal-wallis of sex~size) and calculate colinearity (variance inflation factor, GVIF). To do this, we created a new sex column with three factors: female, male.large, and male.small. That way we won't have to worry about one of our groups (ie females) not having a factor for mating strategy. 

```{r}
# Create swordtail only data set
O.swordtail_1 <- O.datTot[O.datTot$Species =="Swordtail",]

# Create a new sex column based off size categories
O.swordtail_1 <- O.swordtail_1 %>%
  mutate(sex.2 = dplyr::recode(Size.cat,
                        "large" = "male.courtship",
                        "small" = "male.coercion",
                        "reg"   = "female"),
         sex.2 = factor(sex.2, levels = c( "female", "male.courtship", "male.coercion")))

# Correlation test
kruskal.test(Size.mm ~ sex.2, data = O.swordtail_1) #significant (p<0.001)

# Collinearity test
O.mod_test <- lm(CTmax.C ~ sex.2 + Size.mm, data = O.swordtail_1)
vif(O.mod_test) #moderate collinearity for size.mm (3.1); low collinearity for sex.2 (1.75)


### CONCLUSION: adding size would be redundant
```

Since there is only one model (temp ~ sex2), there is no need for model selection. Instead, we will run this model for both temperature categories, check residuals and summarize the models with robust standard errors if necessary.


```{r}
O.SWmod1 <- lm(CTmax.C ~ sex.2, data = O.swordtail_1)
O.SWmod2 <- lm(CTmin.C ~ sex.2, data = O.swordtail_1)

```


### Residual checks

Here, we will check if the model residuals are normal and homoscedastic using the 'DHARMa' package.

**Results:** Residuals for both models (CTmax, CTmin) pass.  

```{r}

# CTmax
simulationOutput <- simulateResiduals(O.SWmod1)

plot(simulationOutput)

plotResiduals(simulationOutput)
plotResiduals(simulationOutput, form = O.swordtail_1$sex.2)

  ### All good!

# CTmin
simulationOutput <- simulateResiduals(O.SWmod2)

plot(simulationOutput)

plotResiduals(simulationOutput)
plotResiduals(simulationOutput, form = O.swordtail_1$sex.2)

  ### All good!


```


### Model Summary

**Results:** Mating strategy did not significantly affect thermal tolerance within swordtails across both temperature categories (CTmax: p=0.119; CTmin: p=0.343). This result is mirrored in all pairwise comparisons. For CTmax, the male-coercion coefficient was significant (p=0.04) but in conjunction with the other results, there is little support for any group differences. *These results mirror full-data results.*

```{r}

# CTmax

 ## Overall strategy effect
Anova(O.SWmod1, type="II")

 ## EMM & Pairwise comparisons

PW.sex.max <- emmeans(O.SWmod1, "sex.2")
summary(PW.sex.max) #EMM
pairs(PW.sex.max) #Pairwise results
      
  ## Plotting EMM

PW.df <- as.data.frame(PW.sex.max) %>%
  mutate(
    sex.2 = case_when(
      sex.2 == "female" ~ "Female",
      sex.2 == "male.courtship" ~ "Male, \ncourtship",
      sex.2 == "male.coercion" ~ "Male, \ncoercion"))


ggplot(PW.df, aes(x=sex.2, y=emmean)) +
  geom_point(size=2) +
  geom_errorbar(aes(ymin=lower.CL, ymax=upper.CL), width=0.1) +
  scale_y_continuous(breaks = scales::pretty_breaks(n = 10)) +
  ylab("Predicted mean temperature (°C) \n") +
  xlab("\n Mating strategy") +
  ggtitle("\n CTmax") +
  theme_classic()

  
  ## Coefficients
summary(O.SWmod1)





# CTmin

## Overall strategy effect
Anova(O.SWmod2, type="II")

 ## EMM & Pairwise comparisons

PW.sex.min <- emmeans(O.SWmod2, "sex.2")
summary(PW.sex.min) #EMM
pairs(PW.sex.min) #Pairwise results
      
  ## Plotting EMM

PW.df <- as.data.frame(PW.sex.min)

ggplot(PW.df, aes(x=sex.2, y=emmean)) +
  geom_point(size=2) +
  geom_errorbar(aes(ymin=lower.CL, ymax=upper.CL), width=0.1) +
  scale_y_continuous(breaks = scales::pretty_breaks(n = 10)) +
  ylab("Predicted mean temperature (°C) \n") +
  xlab("\n Mating strategy") +
  ggtitle("CTmin") +
  theme_classic()
  
  ## Coefficients
summary(O.SWmod2)


```

# **Power Analysis**

Based on the power of our current study, we will calculate the average temperature difference needed to detect a significant sex effect. 

**Results:** The actual temperature difference between males and females in each category was much lower than the simulated temperature differences needed to detect a sex effect given the power of our study (Actual: CTmax=0.3, CTmin=0.2; Simulated: CTmax=0.7, CTmin=0.9). While we found no statistically significant sex effect in any temperature category, smaller biologically meaningful differences can not be ruled out due to our limited sample size.

***NOTE:*** This will increase loading time; to reduce loading time, remove/comment out the following section and keep the comparisons table to view results.

Power simulation:

```{r}
# Function to simulate power and find minimum detectable sex difference
# Inputs:
#   nsim: number of simulations per effect size
#   species: vector of species names
#   nM, nF: vectors of males/females per species
#   sigma: residual SD from linear model
#   alpha: significance threshold
#   power_target: desired power (e.g., 0.8)
#   delta_range: vector of sex differences to test
#   trait_name: name of the trait (for labeling)

simulate_MDD <- function(nsim=1000, species, nM, nF, sigma, alpha=0.05,
                         power_target=0.8, delta_range=seq(0, 3, by=0.1),
                         trait_name="Trait") {
  
  power_results <- data.frame(delta = delta_range, power = NA)
  
  for(i in seq_along(delta_range)){
    sig_count <- 0
    for(j in 1:nsim){
      temp <- c()
      sex <- c()
      sp <- c()
      
      for(k in seq_along(species)){
        # simulate male and female data around delta/2 difference
        temp <- c(temp,
                  rnorm(nM[k], mean = delta_range[i]/2, sd = sigma),
                  rnorm(nF[k], mean = -delta_range[i]/2, sd = sigma))
        sex <- c(sex, rep(c("M","F"), times=c(nM[k], nF[k])))
        sp <- c(sp, rep(species[k], times=nM[k]+nF[k]))
      }
      
      df <- data.frame(temp=temp, sex=sex, species=sp)
      
      # Fit linear model
      m <- lm(temp ~ sex + species, data=df)
      
      # extract p-value for sex
      pval <- summary(m)$coefficients["sexM","Pr(>|t|)"]
      if(!is.na(pval) & pval < alpha) sig_count <- sig_count + 1
    }
    
    power_results$power[i] <- sig_count / nsim
  }
  
  # Find minimum detectable delta (closest to target power)
  closest <- power_results$delta[which.min(abs(power_results$power - power_target))]
  
  cat("\n---", trait_name, "---\n")
  cat("Minimum detectable sex difference for", power_target*100, "% power: ", closest, "\n\n")
  
  return(list(power_curve = power_results, MDD = closest))
} ### END FUNCTION

# ==========================

species <- c("sailfin","mosquitofish","swordtail")
nM <- c(16, 9, 15)
nF <- c(14, 5, 12)

# Replace these sigma values with the residual SD from your real LMs
sigma_CTmax <- 0.6598546   # summary(O.mod1f)$sigma
sigma_CTmin <- 1.350459   # summary(O.mod2f)$sigma

# Run simulations
MDD_CTmax <- simulate_MDD(species=species, nM=nM, nF=nF, sigma=sigma_CTmax, trait_name="CTmax") #0.4°C
MDD_CTmin <- simulate_MDD(species=species, nM=nM, nF=nF, sigma=sigma_CTmin, trait_name="CTmin") #0.9°C

CTmax_df <- MDD_CTmax$power_curve %>% mutate(Trait="CTmax", MDD=MDD_CTmax$MDD)
CTmin_df <- MDD_CTmin$power_curve %>% mutate(Trait="CTmin", MDD=MDD_CTmin$MDD)
power_all <- bind_rows(CTmax_df, CTmin_df)

# Plot
ggplot(power_all, aes(x=delta, y=power, color=Trait)) +
  geom_line(size=1.1) +
  geom_hline(yintercept=0.8, linetype="dashed", color="black") +
  geom_vline(aes(xintercept=MDD, color=Trait), linetype="dotted", size= 1) +
  labs(title="Power curves and minimum detectable sex differences",
       x="Sex difference (°C)",
       y="Power",
       color="Trait") +
  theme_minimal() +
  theme(text = element_text(size=12)) + 
  scale_color_manual(name="Trait",
                    values = c("CTmax"="#E1AF00", "CTmin"="#78B7C5"),
                    labels = c("CTmax"="CTmax", "CTmin"="CTmin"))



```




Comparisons:

```{r}
males <- O.rawTempData[O.rawTempData$Sex == "male",]
females <- O.rawTempData[O.rawTempData$Sex == "female",]

(mean(males$CTmax.C))- (mean(females$CTmax.C)) #-0.137°C
(mean(males$CTmin.C))- (mean(females$CTmin.C)) #0.161°C

df <- data.frame(
  type = c("CTmax", "CTmin"),
  Actual   = c(0.1, 0.2),
  Simulated   = c(0.4, 0.9))

df %>%
  gt() %>%
    cols_label(type ="Temperature Category",
               Actual = "Actual difference (°C)",
               Simulated = "Simulated difference needed (°C)") %>%
  tab_style(
      style = cell_borders(sides = "right", color = "lightgray", weight = px(1.5)),
      locations = cells_body(columns = c(type, Actual))) %>%
  cols_align(align = "center", everything())  %>%
  opt_horizontal_padding(scale=2)

```


# **Summary**

While exact values are slightly different, the removal of outliers ***did not*** alter our overall results. As stated previously, the results reported in the manuscript will reflect those of the full data set.

***

In both trials (CTmax, CTmin), sex did ***not*** influence thermal tolerance. Instead, only species differences affected thermal tolerance. In general, mosquitofish tolerated the most extreme temperatures, followed by sailfin. Swordtails had the narrowest and weakest thermal tolerance, owing likely to the consistency of temperature at their place of origin (e.g spring-fed rivers).

Likewise, mating strategy did ***not*** influence thermal tolerance, further confirming that this is likely no detectable sex-biased thermal differences in our current setup. 

A power analysis shows that the actual detected sex differences in both categories is much lower than the simulated temperature differences needed in order to detect a sex-based effect given the power of our study. Therefore, smaller biologically meaningful differences can not be ruled out due to our limited sample size.


***
***
***
