---
title: "Rerun-outlier-removal"
author: "Allison Davis Connelly & Sophia McKelvey"
date: "2026-04-23"
output: 
  html_document:
    toc: TRUE
    toc_depth: 5
    toc_float:
      collapsed: FALSE
      
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Temperature experiment

**PURPOSE:** Our interest lies in whether sex-baised temperature tolerance exists in 3 livebearing fishes: sailfin mollies (*Poecilia latipinna*), Western mosquitofish (*Gambusia affinis*), and Panuco swordtails (*Xiphophorous nigrensis*). This stems from an observation of males dying when sampling during hot months but females surviving fine. 

**MAIN AIMS:**
  (1) compare female and male thermal tolerance across these three species of livebearing fishes (Family: Poeciliidae)
  (2) compare thermal tolerance across mating strategies within a single livebearing species (*X. nigrensis*)

**HYPOTHESIS:** Preexisting sex differences (e.g. sexually selected traits, size, behavior etc.) decrease resistance to environmental temperature change.

**PREDICTIONS:**
  - Males of fish with elaborate traits have a narrower thermal tolerance--i.e. lower thermal maximums and higher thermal minimums--compared to females of these species.
  - Males of fish without elaborate traits will have a similar tolerance to females of the same species.
  - Males with traits used in courtship (i.e. bright coloration and elongated fins) have a narrower thermal tolerance compared to males that use coercion (within swordtails).
  


***
***
***


# **Data**

All analyses were conducted using R version 4.4.2 

## Load required libraries

Remove the '#' in the code below if it is the first time installing library. 

```{r}
#install.packages("curl")             #v5.2.1
library(curl)

#install.packages("dplyr")            #v1.1.4
library(dplyr)

#install.packages("summarytools")     #v1.1.4
library(summarytools)

#install.packages("gt")               #v0.11.1
library(gt)

#install.packages("tidyr")            #v1.3.1
library(tidyr)
  
#install.packages("MuMIn)              #v1.48.4
library(MuMIn)

#install.packages("lmerTest")         #v3.1.3
library(lmtest)

#install.packages("sandwich")         #v3.1-1
library(sandwich)

#install.packages("ggplot2")          #v3.5.1
library(ggplot2)

#install.packages("ggpubr")          #v0.6.2
library(ggpubr)

#install.packages("DHARMa")           #v0.4.7
library(DHARMa)

#install.packages("emmeans")          #v1.10.5
library(emmeans)

#install.packages("car")              #v3.1-2
library(car)

#install.packages("patchwork")        #v1.3.2
library(patchwork)


```

## Read in data

```{r}

rawTempData <- curl("https://raw.githubusercontent.com/sophia-m0823/TestRStudio/main/Temp%20ID%20-%20Completed%20Data.csv")

rawTempData <- read.csv(rawTempData, header = TRUE, sep = ",", stringsAsFactors = TRUE)
head(rawTempData)

# add common names 
rawTempData$Species <- factor(rawTempData$Species,
                        levels = c("X. nigrensis", "P. latipinna", "G. affinis"),
                        labels = c("Swordtail", "Sailfin", "Mosquitofish"))

# remove mosquitofish ctmax outliers

mosq <- rawTempData[rawTempData$Species == "Mosquitofish",]
hist(mosq$CTmax.C)
(out<- boxplot.stats(mosq$CTmax.C)$out) # three values <35degress
(out_ind <- which(mosq$CTmax.C %in% c(out))) # rows 14-16; Rory, Loreli, Lane

rawTempData <- rawTempData %>%
  filter(!Name %in% c("Rory", "Loreli", "Lane"))

```


## Summery stats 

Trial stats: average CTmax now higher for mosquitofish (38.18 vs 37.58); three less females (9m:5f)

```{r}
#remove unnecessary columns
datTot <- rawTempData[,-c(5:7, 10)]

dat.long <- datTot %>%
  pivot_longer(
    cols = c(CTmax.C, CTmin.C),
    names_to = "Trial",
    values_to = "Value") %>%
  mutate(
    Trial = dplyr::recode(
      Trial,
      "CTmax.C" = "CTmax",
      "CTmin.C" = "CTmin"))

summ.stats <- dat.long %>%
  group_by(Species, Trial) %>%
  summarize(
    mean = mean(Value, na.rm = TRUE),
    sd   = sd(Value, na.rm = TRUE),
    min  = min(Value, na.rm = TRUE),
    med  = median(Value, na.rm = TRUE),
    max  = max(Value, na.rm = TRUE),
    .groups = "drop")


sex.df <- datTot %>%
  count(Species, Sex) %>%
  tidyr::pivot_wider(
    names_from  = Sex,
    values_from = n,
    values_fill = 0)

summ.stats.all <- summ.stats %>%
  left_join(sex.df, by = "Species") %>%
  mutate(
    mean_sd = sprintf("%.2f (%.2f)", mean, sd))


summ.stats.all %>%
  arrange(Species, Trial) %>%
  select(Species, Trial, mean_sd, min, med, max, male, female) %>%
  gt(rowname_col = "Trial", groupname_col = "Species") %>%
  cols_label(
    Species = "Species",
    Trial   = "Trial",
    mean_sd = "Mean (°C) (SD)",
    min     = "Minimum (°C)",
    med     = "Median (°C)",
    max     = "Maximum (°C)",
    male    = "Males",
    female  = "Females") %>%
  fmt_number(
    columns = c(min, med, max),
    decimals = 2) %>%
  tab_style(
      style = cell_borders(sides = "right", color = "lightgray", weight = px(1.5)),
      locations = cells_body(columns = c(mean_sd, min, med, max, male))) %>%
  cols_align("center", everything()) %>%
  opt_horizontal_padding(scale = 2)



```


Size stats: barely any change


```{r}
# Average lengths by species & sex

sizes <- datTot %>%
  group_by(Species, Sex) %>%
  summarise(
    avg.len = mean(Size.mm),
    st.dev = sd(Size.mm),
    .groups = "drop")

 sizes.fmt <- sizes %>%
   mutate(
    size.display = sprintf("%.1f (%.2f)", avg.len, st.dev)) %>%
  select(Species, Sex, size.display) %>%
  pivot_wider(
    names_from = Sex,
    values_from = size.display)

m.f.sizes <- sizes.fmt %>%
  gt() %>%
  tab_header(title = "Average lengths (mm)") %>%
    cols_label(Species ="Species",
               female = "Female (SD)",
               male = "Male (SD)") %>%
  tab_style(
      style = cell_borders(sides = "right", color = "lightgray", weight = px(1.5)),
      locations = cells_body(columns = c(Species, female))) %>%
  tab_style(
    style = cell_text(align = "center"),
    locations = cells_column_labels(columns = everything())) %>%
  opt_horizontal_padding(scale=2)
m.f.sizes

### boxplot
dat.plot <- datTot %>%
  mutate(
    Sex.plot = case_when(
      Species == "Swordtail" & Sex == "male" & Size.cat == "small" ~ "Male (small)",
      Species == "Swordtail" & Sex == "male" & Size.cat == "large" ~ "Male (large)",
      Sex == "male"                                             ~ "Male",
      Sex == "female"                                           ~ "Female"))


ggplot(dat.plot, aes(x = Sex.plot, y = Size.mm)) +
  geom_rect(
    data = subset(datTot, Species == "Sailfin"),
    inherit.aes = FALSE,
    xmin = -Inf, xmax = Inf,
    ymin = -Inf, ymax = Inf,
    fill = "grey95") +
  geom_boxplot() +
  geom_point(color= "#8D8680", size=2, alpha=0.5) +
  facet_wrap(~ Species, scales = "free_x") +
  scale_x_discrete(
  labels = function(x) {
    ifelse(x %in% c("Male (small)", "Male (large)"),
           gsub(" ", "\n", x),
           x)}) +
  theme_classic() + 
  xlab("") +
  ylab("Length (mm)")
  



# Average lengths by sex

sizes2 <- datTot %>%
  group_by(Sex) %>%
  summarise(
    avg.len = mean(Size.mm),
    st.dev = sd(Size.mm),
    .groups = "drop")
 sizes.fmt2 <- sizes2 %>%
   mutate(
    size.display = sprintf("%.1f (%.2f)", avg.len, st.dev)) %>%
  select(Sex, size.display) %>%
  pivot_wider(
    names_from = Sex,
    values_from = size.display)
 
 sizes.fmt2 %>%
  gt() %>%
  tab_header(title = "Average lengths (mm)") %>%
    cols_label(female = "Female (sd)",
               male = "Male (sd)") %>%
  tab_style(
      style = cell_borders(sides = "right", color = "lightgray", weight = px(1.5)),
      locations = cells_body(columns = c(female))) %>%
  tab_style(
    style = cell_text(align = "center"),
    locations = cells_column_labels(columns = everything())) %>%
  opt_horizontal_padding(scale=2)

### boxplot
ggplot(datTot, aes(x = Sex, y = Size.mm)) +
  geom_boxplot() +
  geom_point(color= "#8D8680", size=2, alpha=0.5) +
  theme_classic() +
  xlab("") +
  ylab("Length (mm)") +
  scale_x_discrete(labels=c('Female', 'Male'))
 
 
```


# **Analysis**

## **CTmax**

### Model Selection


**Results:** The model containing just species as a dependent variable (mod1f) is still the most supported in our model selection (ΔAICc=0.0; weight=0.676). The model with both species and sex has a delta value of about 2 now (ΔAICc=2.24; weight=0.221), but we will still keep this model for the purpose of illustrating the limited effect of sex. 

```{r}
# All models
mod1 <- lm(CTmax.C ~ Species*Sex + Size.mm, data = datTot) #full model with interaction effect
mod1b <- lm(CTmax.C ~ Species + Sex + Size.mm, data = datTot) #full model, no interaction
mod1c <- lm(CTmax.C ~ Species + Sex, data = datTot) #no size effect
mod1d <- lm(CTmax.C ~ Sex + Size.mm, data = datTot) #no species effect
mod1e <- lm(CTmax.C ~ Sex, data = datTot) #sex only effect
mod1f <- lm(CTmax.C ~ Species, data = datTot) #species only effect
mod1g <- lm(CTmax.C ~ Size.mm, data = datTot) #size only effect

# Model selection
model.sel(mod1, mod1b, mod1c, mod1d, mod1e, mod1f, mod1g)

  ### Mod1f (species only effect) is the most supported model (ΔAICc=0; weight=0.498)
  ### Mod1c (species and sex, no size) also has a ΔAICc < 2 (ΔAICc=0.7; weight=0.350), so will keep to illustrate the limited effects of sex, but Mod1f is best.

```

### Residual checks


**Results:** The species only model (most supported) NO LONGER shows a significant outlier deviation. Didn't realize those three trials were throwing the tails off so much. The sex and species model also passes tests. Robust standard errors no longer needed.

```{r}

# Species only model
simulationOutput <- simulateResiduals(mod1f)

plot(simulationOutput)

plotResiduals(simulationOutput)
plotResiduals(simulationOutput, form = rawTempData$Species)
   ### All good!

# Species and sex model
simulationOutput2 <- simulateResiduals(mod1c)

plot(simulationOutput2)

plotResiduals(simulationOutput2)
plotResiduals(simulationOutput2, form = rawTempData$Species)
plotResiduals(simulationOutput2, form = rawTempData$Sex)

  ### All good!


```


### Model Summary

**Results:** There is a significant effect of species for CTmax (P<0.001). Pairwise comparisons show a significant difference between swordtails and sailfin (p<0.001) and mosquitofish (p<0.001), but no difference between sailfin & mosquitofish (p=0.687). Sex was confirmed as not a significant predictor in the model (estimate=0.047, SE=0.200, p=0.815).

```{r}

# SPECIES ONLY MODEL

  ## Regular check (not robust)
coeftest(mod1f, vcov. = vcovHC(mod1f, type= "HC3"))

  ## Overall species effect
Anova(mod1f, type="II")

 ## EMM & Pairwise comparisons

PW.sp.max <- emmeans(mod1f, "Species")
summary(PW.sp.max) #EMM
pairs(PW.sp.max) #Pairwise results; now shows significant for all three species
      
  ## Plotting EMM

PW.df <- as.data.frame(PW.sp.max)

df.pvals <- data.frame(
  group1 = c("Swordtail", "Sailfin", "Swordtail"),
  group2 = c("Sailfin", "Mosquitofish", "Mosquitofish"),
  label  = c("p<0.001*", "ns", "p < 0.001*"),
  y.position = c(39, 39, 39))

df.pvals$fontface <- ifelse(df.pvals$label == "ns", "plain", "bold")

emmean.max<-ggplot(PW.df, aes(x=Species, y=emmean)) +
  geom_point(size=2) +
  geom_errorbar(aes(ymin=lower.CL, ymax=upper.CL), width=0.1) +
  scale_y_continuous(breaks = scales::pretty_breaks(n = 10)) +
  stat_pvalue_manual(
    df.pvals,
    label = "label",
    tip.length = 0.01,
    bracket.size = 0.5,
    step.increase = 0.1,
    fontface = "fontface") +
  ylab("Predicted mean temperature (°C) \n") +
  xlab("\n Species") +
  theme_classic()
emmean.max

# SPECIES & SEX MODEL

coeftest(mod1c, vcov. = vcovHC(mod1c, type= "HC3"))


```

## **CTmin**

### Model Selection


**Results:** The full model containing no interactions was the most supported in our model selection (ΔAICc=0.0; weight=0.337). However, the species only model and the species+sex (no size effect) model also have a ΔAICc <2 (species only: ΔAICc= 0.04, weight= 0.329; spp+sex: ΔAICc = 0.35, weight=0.283).

```{r}
# All models
mod2 <- lm(CTmin.C ~ Species*Sex + Size.mm, data = datTot) #full model with interaction effect
mod2b <- lm(CTmin.C ~ Species + Sex + Size.mm, data = datTot) #full model, no interaction
mod2c <- lm(CTmin.C ~ Species + Sex, data = datTot) #no size effect
mod2d <- lm(CTmin.C ~ Sex + Size.mm, data = datTot) #no species effect
mod2e <- lm(CTmin.C ~ Sex, data = datTot) #sex only effect
mod2f <- lm(CTmin.C ~ Species, data = datTot) #species only effect
mod2g <- lm(CTmin.C ~ Size.mm, data = datTot) #size only effect

# Model selection
model.sel(mod2, mod2b, mod2c, mod2d, mod2e, mod2f, mod2g)

  ### Mod2b (full model, no interactions) is the most supported model (ΔAICc=0; weight=0.337)
  ### Mod2f (species only) also has a ΔAICc < 2 (ΔAICc=0.04; weight=0.329)
  ### Mod2c (species + sex) also has a ΔAICc < 2 (ΔAICc=0.35; weight=0.3283)
  ### Will check out the summaries below to see if sex or size truly impacts the model; if so, will average the models

```

### Residual checks

**Results:** Both the full model and species only are heteroscedastic. This is driven by species-level differences in variation. Robust standard errors will allow us to assess the significance of the effects despite this heteroscedasticity. Species + sex model passes residual checks, although does show a failed Levene's test when plotted by species (not sure if that means I do normal or robust SEs).

```{r}

# full model, no interactions
simulationOutput <- simulateResiduals(mod2b)

plot(simulationOutput)

plotResiduals(simulationOutput)
plotResiduals(simulationOutput, form = rawTempData$Species)

  ### Failed predicted vs quantile deviations & Levene's test


# Species only model
simulationOutput <- simulateResiduals(mod2f)

plot(simulationOutput)

plotResiduals(simulationOutput)
plotResiduals(simulationOutput, form = rawTempData$Species)

  ### Failed Levene's test driven by species-level variation differences


# Species and sex model
simulationOutput2 <- simulateResiduals(mod2c)

plot(simulationOutput2)

plotResiduals(simulationOutput2)
plotResiduals(simulationOutput2, form = rawTempData$Species)
plotResiduals(simulationOutput2, form = rawTempData$Sex)

  ### Fine in overall plots, but by species shows a failure in Levene's test


```


### Model Summary

Did a quick summary of the top three models: species is the only significant effect in all three models (sex and size aren't significant). Will stick with the species only model as the top supported and then show the species + sex model just to highlight the effects of sex.

**Results:** There is a significant effect of species for CTmin (P<0.001). Pairwise comparisons show all species as significantly different: swordtail-sailfin= p<0.001; swordtail-mosquitofish=p<0.001; sailfin-mosquitofish= p=0.001). Sex was a signifcant predictor in the sex + species model (estimate=0.444, SE=0.200, p=0.030). Again, the model selection picked it out as a competitive model, but the summary doesn't show an effect of sex so really not sure how to interpret this.  

```{r}
# QUICK CHECK
summary(mod2b) #sex and size not significant
summary(mod2f) #species still significant
summary(mod2c) #size not significant

  ### Will proceed with mod2f (species only) as this is the most parsimonious, and adding sex & size doesn't add much

# SPECIES ONLY MODEL

 ## Robust variance-covariance matrix
robVCV.sp <- vcovHC(mod2f, type= "HC3")

 ## Overall species effect
Anova(mod2f, vcov = robVCV.sp, type="II")

 ## EMM & Pairwise comparisons

PW.sp.min <- emmeans(mod2f, "Species", vcov. = robVCV.sp)
summary(PW.sp.min) #EMM
pairs(PW.sp.min) #Pairwise results
      
  ## Plotting EMM

PW.df <- as.data.frame(PW.sp.min)

df.pvals <- data.frame(
  group1 = c("Swordtail", "Sailfin", "Swordtail"),
  group2 = c("Sailfin", "Mosquitofish", "Mosquitofish"),
  label  = c("p<0.001*", "p = 0.004*", "p<0.001*"),
  y.position = c(10.5, 10.5, 10.5))

df.pvals$fontface <- ifelse(df.pvals$label == "ns", "plain", "bold")

emmean.min <- ggplot(PW.df, aes(x=Species, y=emmean)) +
  geom_point(size=2) +
  geom_errorbar(aes(ymin=lower.CL, ymax=upper.CL), width=0.1) +
  scale_y_continuous(breaks = scales::pretty_breaks(n = 8)) +
  ylab("Predicted mean temperature (°C) \n") +
  stat_pvalue_manual(
    df.pvals,
    label = "label",
    tip.length = 0.01,
    bracket.size = 0.5,
    step.increase = 0.1,
    fontface = "fontface") +
  xlab("\n Species") +
  theme_classic()
emmean.min

# SPECIES & SEX MODEL

coeftest(mod2c, vcov. = vcovHC(mod1c, type= "HC3"))


```

## **Across mating strategies**

The second aim in our analysis is to compare thermal tolerance across male mating strategies within one species (Panuco swordtail). We used swordtail males of two categories: large, courting males (standard length >32mm) or small, coercive males (standard length <25mm with clearly defined gonopodium). If preexisting sex-based differences (e.g. bright coloration, elongated fins, courtship) affect thermal tolerance, we predict large males to have a lower thermal tolerance than small males, and small males to mirror the thermal tolerance of females.  

### Models

Since male mating strategy is associated with size, adding size to our models would be redundant. Just to confirm, we will run a correlation test (kruskal-wallis of sex~size) and calculate colinearity (variance inflation factor, GVIF). To do this, we created a new sex column with three factors: female, male.large, and male.small. That way we won't have to worry about one of our groups (ie females) not having a factor for mating strategy. 

```{r}
# Create swordtail only data set
swordtail_1 <- datTot[datTot$Species =="Swordtail",]

# Create a new sex column based off size categories
swordtail_1 <- swordtail_1 %>%
  mutate(sex.2 = dplyr::recode(Size.cat,
                        "large" = "male.courtship",
                        "small" = "male.coercion",
                        "reg"   = "female"),
         sex.2 = factor(sex.2, levels = c( "female", "male.courtship", "male.coercion")))

# Correlation test
kruskal.test(Size.mm ~ sex.2, data = swordtail_1) #significant (p<0.001)

# Collinearity test
mod_test <- lm(CTmax.C ~ sex.2 + Size.mm, data = swordtail_1)
vif(mod_test) #moderate collinearity for size.mm (3.0); low collinearity for sex.2 (1.73)


### CONCLUSION: adding size would be redundant
```

Since there is only one model (temp ~ sex2), there is no need for model selection. Instead, we will run this model for both temperature categories, check residuals and summarize the models with robust standard errors if necessary.


```{r}
SWmod1 <- lm(CTmax.C ~ sex.2, data = swordtail_1)
SWmod2 <- lm(CTmin.C ~ sex.2, data = swordtail_1)

```


### Residual checks

Here, we will check if the model residuals are normal and homoscedastic using the 'DHARMa' package.

**Results:** Residuals for both models (CTmax, CTmin) pass.  

```{r}

# CTmax
simulationOutput <- simulateResiduals(SWmod1)

plot(simulationOutput)

plotResiduals(simulationOutput)
plotResiduals(simulationOutput, form = swordtail_1$sex.2)

  ### All good!

# CTmin
simulationOutput <- simulateResiduals(SWmod2)

plot(simulationOutput)

plotResiduals(simulationOutput)
plotResiduals(simulationOutput, form = swordtail_1$sex.2)

  ### All good!


```


### Model Summary

All models pass residual checks, so we will determine the overall effect of mating strategy, calculate the estimated marginal means, and run a pairwise strategy comparison without an additional variance-covariance specification.

**Results:** Mating strategy did not significantly affect thermal tolerance within swordtails across both temperature categories (CTmax: p=0.647; CTmin: p=0.378). This result is mirrored in all pairwise comparisons. 

```{r}

# CTmax

 ## Overall strategy effect
Anova(SWmod1, type="II")

 ## EMM & Pairwise comparisons

PW.sex.max <- emmeans(SWmod1, "sex.2")
summary(PW.sex.max) #EMM
pairs(PW.sex.max) #Pairwise results
      
  ## Plotting EMM

PW.df <- as.data.frame(PW.sex.max) %>%
  mutate(
    sex.2 = case_when(
      sex.2 == "female" ~ "Female",
      sex.2 == "male.courtship" ~ "Male, \ncourtship",
      sex.2 == "male.coercion" ~ "Male, \ncoercion"))


em.max <- ggplot(PW.df, aes(x=sex.2, y=emmean)) +
  geom_point(size=2) +
  geom_errorbar(aes(ymin=lower.CL, ymax=upper.CL), width=0.1) +
  scale_y_continuous(breaks = scales::pretty_breaks(n = 10)) +
  ylab("Predicted mean temperature (°C) \n") +
  xlab("\n Mating strategy") +
  ggtitle("\n CTmax") +
  theme_classic()
em.max
  
  ## Coefficients
summary(SWmod1)



# CTmin

## Overall strategy effect
Anova(SWmod2, type="II")

 ## EMM & Pairwise comparisons

PW.sex.min <- emmeans(SWmod2, "sex.2")
summary(PW.sex.min) #EMM
pairs(PW.sex.min) #Pairwise results
      
  ## Plotting EMM

PW.df <- as.data.frame(PW.sex.min)

em.min <- ggplot(PW.df, aes(x=sex.2, y=emmean)) +
  geom_point(size=2) +
  geom_errorbar(aes(ymin=lower.CL, ymax=upper.CL), width=0.1) +
  scale_y_continuous(breaks = scales::pretty_breaks(n = 10)) +
  ylab("Predicted mean temperature (°C) \n") +
  xlab("\n Mating strategy") +
  ggtitle("CTmin") +
  theme_classic()
em.min
  
  ## Coefficients
summary(SWmod2)


```



# **Power Analysis**

Based on the power of our current study, we will calculate the average temperature difference needed to detect a significant sex effect. 

**Results:** The actual temperature difference between males and females in each category was much lower than the simulated temperature differences needed to detect a sex effect given the power of our study (Actual: CTmax=0.3, CTmin=0.2; Simulated: CTmax=0.7, CTmin=0.9). While we found no statistically significant sex effect in any temperature category, smaller biologically meaningful differences can not be ruled out due to our limited sample size.

Power simulation:

```{r}
# Function to simulate power and find minimum detectable sex difference
# Inputs:
#   nsim: number of simulations per effect size
#   species: vector of species names
#   nM, nF: vectors of males/females per species
#   sigma: residual SD from linear model
#   alpha: significance threshold
#   power_target: desired power (e.g., 0.8)
#   delta_range: vector of sex differences to test
#   trait_name: name of the trait (for labeling)

simulate_MDD <- function(nsim=1000, species, nM, nF, sigma, alpha=0.05,
                         power_target=0.8, delta_range=seq(0, 3, by=0.1),
                         trait_name="Trait") {
  
  power_results <- data.frame(delta = delta_range, power = NA)
  
  for(i in seq_along(delta_range)){
    sig_count <- 0
    for(j in 1:nsim){
      temp <- c()
      sex <- c()
      sp <- c()
      
      for(k in seq_along(species)){
        # simulate male and female data around delta/2 difference
        temp <- c(temp,
                  rnorm(nM[k], mean = delta_range[i]/2, sd = sigma),
                  rnorm(nF[k], mean = -delta_range[i]/2, sd = sigma))
        sex <- c(sex, rep(c("M","F"), times=c(nM[k], nF[k])))
        sp <- c(sp, rep(species[k], times=nM[k]+nF[k]))
      }
      
      df <- data.frame(temp=temp, sex=sex, species=sp)
      
      # Fit linear model
      m <- lm(temp ~ sex + species, data=df)
      
      # extract p-value for sex
      pval <- summary(m)$coefficients["sexM","Pr(>|t|)"]
      if(!is.na(pval) & pval < alpha) sig_count <- sig_count + 1
    }
    
    power_results$power[i] <- sig_count / nsim
  }
  
  # Find minimum detectable delta (closest to target power)
  closest <- power_results$delta[which.min(abs(power_results$power - power_target))]
  
  cat("\n---", trait_name, "---\n")
  cat("Minimum detectable sex difference for", power_target*100, "% power: ", closest, "\n\n")
  
  return(list(power_curve = power_results, MDD = closest))
} ### END FUNCTION

# ==========================

species <- c("sailfin","mosquitofish","swordtail")
nM <- c(16, 9, 15)
nF <- c(16, 5, 15)

# Replace these sigma values with the residual SD from your real LMs
sigma_CTmax <- 0.8328148   # summary(mod1f)$sigma
sigma_CTmin <- 1.389272   # summary(mod2f)$sigma

# Run simulations
MDD_CTmax <- simulate_MDD(species=species, nM=nM, nF=nF, sigma=sigma_CTmax, trait_name="CTmax") #0.5°C
MDD_CTmin <- simulate_MDD(species=species, nM=nM, nF=nF, sigma=sigma_CTmin, trait_name="CTmin") #0.9°C

CTmax_df <- MDD_CTmax$power_curve %>% mutate(Trait="CTmax", MDD=MDD_CTmax$MDD)
CTmin_df <- MDD_CTmin$power_curve %>% mutate(Trait="CTmin", MDD=MDD_CTmin$MDD)
power_all <- bind_rows(CTmax_df, CTmin_df)

# Plot
ggplot(power_all, aes(x=delta, y=power, color=Trait)) +
  geom_line(size=1.1) +
  geom_hline(yintercept=0.8, linetype="dashed", color="black") +
  geom_vline(aes(xintercept=MDD, color=Trait), linetype="dotted", size= 1) +
  labs(title="Power curves and minimum detectable sex differences",
       x="Sex difference (°C)",
       y="Power",
       color="Trait") +
  theme_minimal() +
  theme(text = element_text(size=12)) + 
  scale_color_manual(name="Trait",
                    values = c("CTmax"="#E1AF00", "CTmin"="#78B7C5"),
                    labels = c("CTmax"="CTmax", "CTmin"="CTmin"))



```


Comparisons:

```{r}
males <- rawTempData[rawTempData$Sex == "male",]
females <- rawTempData[rawTempData$Sex == "female",]

(mean(males$CTmax.C))- (mean(females$CTmax.C)) #0.196°C
(mean(males$CTmin.C))- (mean(females$CTmin.C)) #0.089°C

df <- data.frame(
  type = c("CTmax", "CTmin"),
  Actual   = c(0.3, 0.1),
  Simulated   = c(0.5, 0.9))

df %>%
  gt() %>%
    cols_label(type ="Temperature Category",
               Actual = "Actual difference (°C)",
               Simulated = "Simulated difference needed (°C)") %>%
  tab_style(
      style = cell_borders(sides = "right", color = "lightgray", weight = px(1.5)),
      locations = cells_body(columns = c(type, Actual))) %>%
  cols_align(align = "center", everything())  %>%
  opt_horizontal_padding(scale=2)

```


# **Summary**

WILL UPDATE AFTER DISCUSSING WITH MIKE

In both trials (CTmax, CTmin), sex did ***not*** influence thermal tolerance. Instead, only species differences affected thermal tolerance. In general, mosquitofish tolerated the most extreme temperatures, followed by sailfin. Swordtails had the narrowest and weakest thermal tolerance, owing likely to the consistency of temperature at their place of origin (e.g spring-fed rivers).

Likewise, mating strategy did ***not*** influence thermal tolerance, further confirming that this is likely no detectable sex-biased thermal differences in our current setup. 

A power analysis shows that the actual detected sex differences in both categories is much lower than the simulated temperature differences needed in order to detect a sex-based effect given the power of our study. Therefore, smaller biologically meaningful differences can not be ruled out due to our limited sample size.


***
***
***

# Alternative tables & figures

This section will have tables and figures used for publication.

(1) Combined pairwise comparisons- across species

```{r}
pw.ctmax <- pairs(PW.sp.max) %>%
  as.data.frame() %>%
  mutate(
    Trial = "CTmax",
    est_se = sprintf("%.2f (%.2f)", estimate, SE)) %>%
  select(Trial, contrast, est_se, p.value)

pw.ctmin <- pairs(PW.sp.min) %>%
  as.data.frame() %>%
  mutate(
    Trial = "CTmin",
    est_se = sprintf("%.2f (%.2f)", estimate, SE)) %>%
  select(Trial, contrast, est_se, p.value)


pw.all <- dplyr::bind_rows(pw.ctmax, pw.ctmin) %>%
  mutate(
    p.display = case_when(
      p.value < 0.001 ~ "*p* < 0.001",
      p.value < 0.05  ~ paste0(sprintf("%.3f", p.value)),
      TRUE            ~ sprintf("%.3f", p.value)))


pw.all %>%
  gt(groupname_col = "Trial") %>%
  cols_hide(p.value) %>%
  cols_label(
    contrast = "Species contrast",
    est_se  = "Estimate (SE)",
    p.display = md("*p*-value")) %>%
  fmt_markdown(columns = p.display) %>%
  tab_style(
      style = cell_borders(sides = "right", color = "lightgray", weight = px(1.5)),
      locations = cells_body(columns = c(contrast, est_se))) %>%
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_body(
      columns = p.display,
      rows = p.value < 0.05)) %>%
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_row_groups()) %>%
  cols_align("center", everything()) %>%
  opt_horizontal_padding(scale = 2)


```


(2) Combined pairwise comparisons- across mating strategies

```{r}
pw.sex.max <- pairs(PW.sex.max) %>%
  as.data.frame() %>%
  mutate(
    Trial = "CTmax",
    est_se = sprintf("%.2f (%.2f)", estimate, SE)) %>%
  select(Trial, contrast, est_se, p.value)

pw.sex.min <- pairs(PW.sex.min) %>%
  as.data.frame() %>%
  mutate(
    Trial = "CTmin",
    est_se = sprintf("%.2f (%.2f)", estimate, SE)) %>%
  select(Trial, contrast, est_se, p.value)


pw.sex.all <- dplyr::bind_rows(pw.sex.max, pw.sex.min) %>%
  mutate(
    p.display = case_when(
      p.value < 0.001 ~ "*p* < 0.001",
      p.value < 0.05  ~ paste0(sprintf("%.3f", p.value)),
      TRUE            ~ sprintf("%.3f", p.value)))


pw.sex.all %>%
  gt(groupname_col = "Trial") %>%
  cols_hide(p.value) %>%
  cols_label(
    contrast = "Strategies contrast",
    est_se  = "Estimate (SE)",
    p.display = md("*p*-value")) %>%
  fmt_markdown(columns = p.display) %>%
  tab_style(
      style = cell_borders(sides = "right", color = "lightgray", weight = px(1.5)),
      locations = cells_body(columns = c(contrast, est_se))) %>%
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_body(
      columns = p.display,
      rows = p.value < 0.05)) %>%
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_row_groups()) %>%
  cols_align("center", everything()) %>%
  opt_horizontal_padding(scale = 2)


```


(3) Box plots by species by sex

```{r, fig.width=10, fig.height=10}
# CTmax

dat.plot <- dat.plot %>%
  mutate(Sex.plot = case_when(
      Sex == "male" ~ "Male",
      Sex == "female" ~ "Female"))

range.max<-ggplot(dat.plot, aes(x = Sex.plot, y = CTmax.C)) +
  geom_rect(
    data = subset(datTot, Species == "Sailfin"),
    inherit.aes = FALSE,
    xmin = -Inf, xmax = Inf,
    ymin = -Inf, ymax = Inf,
    fill = "grey95") +
  geom_boxplot() +
  geom_point(color= "#8D8680", size=2, alpha=0.5) +
  facet_wrap(~ Species, scales = "free_x") +
  theme_classic() + 
  xlab("Sex") +
  ylab("Temperature at LOE (°C)")
range.max
  

# CTmin

range.min <- ggplot(dat.plot, aes(x = Sex.plot, y = CTmin.C)) +
  geom_rect(
    data = subset(datTot, Species == "Sailfin"),
    inherit.aes = FALSE,
    xmin = -Inf, xmax = Inf,
    ymin = -Inf, ymax = Inf,
    fill = "grey95") +
  geom_boxplot() +
  geom_point(color= "#8D8680", size=2, alpha=0.5) +
  facet_wrap(~ Species, scales = "free_x") +
  theme_classic() + 
  xlab("Sex") +
  ylab("Temperature at LOE (°C)")
range.min
  
range.spp <- (range.max + range.min + plot_layout(axis_titles = "collect"))

em.max.spp <- emmean.max + xlab("") + ggtitle("CTmax") + ylim(NA, 41)
em.min.spp <- emmean.min + xlab("") + ggtitle ("CTmin") + ylim(NA, 13)
em.spp <- (em.max.spp + em.min.spp + plot_layout(axis_titles = "collect"))


((em.spp/range.spp) + plot_annotation(tag_levels = "A"))

```


(4) Box plots by species by sex category

```{r, fig.width=7, fig.height=5}
dat.plot.sex <- swordtail_1 %>%
  mutate(
    Sex.plot = case_when(
      sex.2 == "male.coercion" ~ "Male (coercion)",
      sex.2 == "male.courtship" ~ "Male (courtship)",
      sex.2 == "female" ~ "Female")) 

# CTmax

max.plot <- ggplot(dat.plot.sex, aes(x = Sex.plot, y = CTmax.C)) +
  geom_boxplot() +
  geom_point(color= "#8D8680", size=2, alpha=0.5) +
  scale_x_discrete(
  labels = function(x) {
    ifelse(x %in% c("Male (coercion)", "Male (courtship)"),
           gsub(" ", "\n", x),
           x)}) +
  theme_classic() + 
  xlab("\n Mating strategy") +
  ylab("Temperature at LOE (°C)")
  
# CTmin

min.plot <- ggplot(dat.plot.sex, aes(x = Sex.plot, y = CTmin.C)) +
  geom_boxplot() +
  geom_point(color= "#8D8680", size=2, alpha=0.5) +
  scale_x_discrete(
  labels = function(x) {
    ifelse(x %in% c("Male (coercion)", "Male (courtship)"),
           gsub(" ", "\n", x),
           x)}) +
  theme_classic() + 
  xlab("\n Mating strategy") +
  ylab("Temperature at LOE (°C)")


 
 (max.plot + min.plot + plot_layout(axis_titles = "collect"))
range.swd <-  (max.plot + min.plot + plot_layout(axis_titles = "collect"))

```

(5) Combined emmeans for swordtails

```{r, fig.width=7, fig.height=5}
my.labels <- c("Female", "Male, \n courtship", "Male, \n coercion")
em.max <- em.max + scale_x_discrete(labels = my.labels) + xlab("")
em.min <- em.min + scale_x_discrete(labels = my.labels) + xlab("")

(em.max + em.min + plot_layout(axis_titles = "collect"))

em.swd <- (em.max + em.min + plot_layout(axis_titles = "collect"))

```

```{r, fig.width=10, fig.height=10}

((em.swd/range.swd) + plot_annotation(tag_levels = "A"))


```

(6) Simple test comparing male and female lengths

```{r}
normal <- datTot %>%
  group_by(Sex, Species) %>%
summarize(statistic =shapiro.test(Size.mm)$statistic,
          p.value = shapiro.test(Size.mm)$p.value)

normal #all normal except male and female swordtail

normal2 <- datTot %>%
  group_by(Sex) %>%
summarize(statistic =shapiro.test(Size.mm)$statistic,
          p.value = shapiro.test(Size.mm)$p.value)

normal2 #not normal for either sex

size.test <- datTot %>%
  group_by(Species) %>%
  summarize(stat = wilcox.test(Size.mm ~ Sex)$statistic,
            p_value = wilcox.test(Size.mm~Sex)$p.value)

size.test #not significant for swordtail and sailfin, but significant for mosquitofish; this may be due to the low sample size for mosquiofish compared to the other two species or the vastly different sizes of males and females in this species

eff.size.tbl <- datTot %>%
  group_by(Species) %>%
  summarise({
    cd <- effsize::cliff.delta(Size.mm ~ Sex)
    tibble(
      delta= cd$estimate,
    CI_low = cd$conf.int[1],
    CI_high = cd$conf.int[2])})
eff.size.tbl


### minimal effect size for swordtail and sailfin, but large effect size for mosquitofish. This is understandable seeing as females are often MUCH larger than males. However, size still had very little effect on temperature tolerance despite a size difference between sexes in this species. 
```

(7) Lengths figure and table

```{r}
### table

stats_tbl <- eff.size.tbl %>%
  left_join(size.test, by = "Species") %>%
  mutate(
    cliffs_delta = sprintf("%.2f", delta),
    p.display = case_when(
    p_value < 0.001 ~ "***p*** **<** **0.001**",
    p_value < 0.05  ~ paste0("*p* = ", sprintf("%.3f", p_value)),
    TRUE            ~ paste0("*p* = ", sprintf("%.3f", p_value))),

  wilcox_res = paste0(p.display, " (W = ", sprintf("%.0f", stat), ")")) %>%
  select(Species, cliffs_delta, wilcox_res)

size.table <- sizes.fmt %>%
  left_join(stats_tbl, by="Species")

size.table %>%
  gt() %>%
  tab_header(title = "Average lengths (mm)") %>%
  cols_label(
    female       = "Female (SD)",
    male         = "Male (SD)",
    cliffs_delta = "Cliff’s δ",
    wilcox_res   = "Wilcoxon") %>%
  fmt_markdown(columns = wilcox_res) %>%
  tab_style(
      style = cell_borders(sides = "right", color = "lightgray", weight = px(1.5)),
    locations = cells_body(columns = c(Species, female, male, cliffs_delta))) %>%
  tab_style(
    style = cell_text(align = "center"),
    locations = cells_column_labels(columns = everything())) %>%
  opt_horizontal_padding(scale = 2)


### boxplot

df.pval <- data.frame(label = c("ns", "ns", "p<0.001*"),
                      Species = c("Swordtail", "Sailfin", "Mosquitofish"),
                      x = c(1.5, 1.5, 1.5),
                      y = c(39, 39, 39))

bracket_data <- data.frame(
  Species = "Mosquitofish", 
  x = c(1, 1, 2, 2),
  y = c(37.5, 38, 38, 37.5))


dat.plot2 <- datTot %>%
  mutate(
    Sex.plot2 = case_when(
      Sex == "male" ~ "Male",
      Sex == "female" ~ "Female"))


ggplot(dat.plot2, aes(x = Sex.plot2, y = Size.mm)) +
  geom_rect(
    data = subset(datTot, Species == "Sailfin"),
    inherit.aes = FALSE,
    xmin = -Inf, xmax = Inf,
    ymin = -Inf, ymax = Inf,
    fill = "grey95") +
  geom_boxplot() +
  geom_point(color= "#8D8680", size=2, alpha=0.5) +
  facet_wrap(~ Species, scales = "free_x") +
  geom_text(df.pval, mapping = aes(x=x, y=y, label=label))+
  geom_path(
    data = bracket_data, 
    aes(x = x, y = y), 
    inherit.aes = FALSE, 
    linewidth = 0.5) +
  theme_classic() + 
  xlab("") +
  ylab("Length (mm)")
  

```


(8) Temp ranges for all species

```{r}

dat.plot3 <- rawTempData %>%
  mutate(
    Sex.plot3 = case_when(
      Sex == "male" ~ "Male",
      Sex == "female" ~ "Female"))

ggplot(dat.plot3, aes(x = Sex.plot3, y = T.tot)) +
  geom_rect(
    data = subset(datTot, Species == "Sailfin"),
    inherit.aes = FALSE,
    xmin = -Inf, xmax = Inf,
    ymin = -Inf, ymax = Inf,
    fill = "grey95") +
  geom_boxplot() +
  geom_point(color= "#8D8680", size=2, alpha=0.5) +
  facet_wrap(~ Species, scales = "free_x") +
  theme_classic() + 
  xlab("") +
  ylab("Temperature range (°C) \n(CTmax-CTmin)")
  

```

(9) Paired points

```{r}
datTot2 <- rawTempData[,-c(6:7, 10)]


df_wide <- datTot2 %>%
  rename(Min = CTmin.C, Max = CTmax.C)

sex_colors <- c("male" = "#0072B2", "female" = "#D55E00")
plot_species_dynamic <- function(df_sub) {
  df_sub <- df_sub %>% mutate(Sex = factor(Sex))
  # Calculate min/max for this specific species to define the dynamic scale
  min_val <- min(df_sub$Min, na.rm = TRUE)
  max_val <- max(df_sub$Max, na.rm = TRUE)
  range_min <- max(df_sub$Min, na.rm = TRUE) - min(df_sub$Min, na.rm = TRUE)
  range_max <- max(df_sub$Max, na.rm = TRUE) - min(df_sub$Max, na.rm = TRUE)
  
  # Dynamic transformation
  scale_ratio <- range_max / range_min
  min_to_max <- function(x) (x - min_val) * (range_max / range_min) + min(df_sub$Max)
  max_to_min <- function(x) (x - min(df_sub$Max)) * (range_min / range_max) + min_val 
  # Create jitter for X-axis (the same for both points and segments)
  set.seed(123) # Ensures the jitter is the same every time you run it
  df_sub$x_jitter <- runif(nrow(df_sub), -0.03, 0.03)
  
  ggplot(df_sub, aes(group = Name)) +
    # Draw segments: X is 1+jitter, Xend is 1.5+jitter (tightened spacing)
    geom_segment(aes(x = 1 + x_jitter, xend = 1.2 + x_jitter, 
                     y = Max, yend = min_to_max(Min)), 
                 color = "gray24", alpha = 0.5) +
    # Points
    geom_point(aes(x = 1 + x_jitter, y = Max, shape = Sex, color=Sex), size = 2.5, alpha=0.8) +
    geom_point(aes(x = 1.2 + x_jitter, y = min_to_max(Min), shape = Sex, color=Sex), size = 2.5, alpha=0.8) +
    scale_color_manual(values = sex_colors) +
    scale_shape_manual(values = c(16, 17)) + # 16=Circle, 17=Triangle
    scale_y_continuous(
      name = "CTmax (°C)",
      sec.axis = sec_axis(~max_to_min(.), name = "CTmin (°C)")
    ) +
    # Set X-axis to match the new 1 and 1.5 positions
    scale_x_continuous(breaks = c(1, 1.2), labels = c("CTmax", "CTmin"), limits = c(0.9, 1.3), expand = c(0.01, 0.01)) +
    labs(title = unique(df_sub$Species)) +
    theme_classic() +
    theme(axis.title.x = element_blank())
}

# 2. Apply to each species
p_sword <- plot_species_dynamic(df_wide %>% filter(Species == "Swordtail"))
p_sail  <- plot_species_dynamic(df_wide %>% filter(Species == "Sailfin"))
p_mosq  <- plot_species_dynamic(df_wide %>% filter(Species == "Mosquitofish"))

# 3. Stack them
(p_sword / p_sail / p_mosq) + plot_layout(guides = "collect")


```
